### 2015-02-26 paper/futurework.tex
corrections in related work and future work

Until recently, Rust's closures have been in flux. As of this writing, Rust supports
both stack and heap closures and also supports different manners of access to the environment
(owning, immutable refernce, or mutable reference). In the past, closures have been a
common source of borrow checker bugs, which likely means that proving soundness will be
difficult. However, it also means that it is one of the most essential features to add to Patina.
It would likely be profitable to look to Cyclone \cite{cycregions}, which took great
care to ensure references could not escape via closures.

Traits, which are Rust's version of typeclasses or interfaces, may have similar issues.
Traits can be used to create what is effectively an object.
Since objects and closures are so similar, it is likely that proving trait objects sound
will face similar difficulties.

>>>>

Until recently, Rust's closures have been in flux. As of this writing, Rust supports
both stack and heap closures and also supports different manners of access to the environment
(owning, immutable refernce, or mutable reference). In the past, closures have been a
common source of borrow checker bugs, which likely means that proving soundness will be
difficult. However, it also means that it is one of the most essential features to add to Patina.
It would likely be profitable to look to Cyclone \cite{cycregions}, which took great
care to ensure references could not escape via closures or existential types.

Traits, which are Rust's version of typeclasses or interfaces, may have similar issues.
Traits can be used to create what is effectively an object or an existential type.
Since objects and closures are so similar, it is likely that proving trait objects sound
will face similar difficulties. Rust's closures are actually implemented as syntactic sugar
for certain trait implementations in the current version. 
It is unclear whether this means formalization of closures will be easier
or whether formalization of traits will be harder.

### 2015-02-26 paper/relatedwork.tex
corrections in related work and future work

Both ML Kit \cite{mlkit-4.3.0, mlkitregions}
and Cyclone \cite{cycregions} heavily influenced the design of Rust.
However, both of those languages, along with most research in the area,
uses regions for memory management. In constrast, Rust's memory is managed
by affine types (unique ownership), and its use of regions is for memory
safety. Cyclone's region subtyping is the source of Rust's lifetime subtyping
and so they ensure references do not escape in a similar fashion.
Unlike Rust, Cyclone does not prevent access to uninitialized memory,
e.g. it is possible to dereference a null pointer.
Unlike Cyclone, Rust does not have first class regions, but Rust also
lacks Cyclone's dynamic regions, which lessens the need for first class regions.

Cyclone later added unique pointers and a notion of borrowing. \cite{cycsafe, cycexperience}
Cyclone's version does not prevent memory leaks, which are possible both from
freeing a unique pointer that contains another live unique pointer and from
overwriting a live unique pointer. Rust statically guarantees no memory leaks in both
of these situations. Cyclone requires an explicit swap operation for using unique versions
of pointers that do not admit null values. Rust will allow any of its pointers to be null,
but guarantees that they will not be used while null. Rust also does not require explicit swaps;
Patina uses them, but that is an artifact of the model.
Cyclone considers unique pointers "consumed" while a borrow exists, 
whereas Rust will allow the borrowed unique pointer to be used in certain situations
(namely, reading through an immutably borrowed unique pointer is allowed).
Cyclone supports abstraction over the aliasability and kind (box versus region) of a pointer,
but Rust does not currently support any such abstraction over pointer types.

@@

Boyland's alias burying \cite{aliasburying} has a similar goal to Rust, i.e.
using static checks rather than destructive reads, but differs in many details.
Boyland a strong version of the uniqueness invariant (true uniqueness) and a weak
version (allowing borrowed aliases), and then develops a system based on maintaing
the strong invariant. Rust meets only the weak uniqueness invariant, but it also
guarantees that aliased, mutable memory never exists, which allows for similar control
over mutation as the strong uniqueness invariant. The general idea of alias burying
is to allow for aliases, but to invalidate them when a unique field is read.
Rust ensures at a unique value will not be read if (mutable) aliases exist.
Rust's borrowed references are also more first-class than Boyland's, which cannot
be stored in data structures or returned from functions.

>>>>

Safe dialects of C, unique pointers, borrowed references, 
regions, and other topics relevant to Rust each have long
and full histories. Rather than attempt to catalogue them in detail
and compare Rust with each of its ancestors, we will limit our discussion
to the most closely related work we are aware of.

Both ML Kit \cite{mlkit-4.3.0, mlkitregions}
and Cyclone \cite{cycregions} heavily influenced the design of Rust.
However, both of those languages, along with most research in the area,
uses regions for memory management. In constrast, Rust's memory is managed
by affine types (unique ownership), and uses regions (lifetimes in Rust)
more for ensuring the safety of borrowed references than for managing memory.
Cyclone's region subtyping is the source of Rust's lifetime subtyping
and so they ensure references do not escape in a similar fashion.
Like Cyclone, Rust prevents access to uninitialized memory,
but Rust goes further than Cyclone by statically preventing dereferences of null pointers.
Unlike Cyclone, Rust does not have first class regions, but Rust also
lacks Cyclone's dynamic regions, which lessens the need for first class regions.

Cyclone later added unique pointers and a notion of borrowing. \cite{cycsafe, cycexperience}
Cyclone's version does not prevent memory leaks, which are possible both from
freeing a unique pointer that contains another live unique pointer and from
overwriting a live unique pointer. Rust statically guarantees no memory leaks in both
of these situations. Cyclone requires an explicit swap operation for using unique versions
of pointers that do not admit null values. Rust will allow any of its pointers to be null,
but guarantees that they will not be used while null. Rust also does not require explicit swaps;
Patina uses them, but that is an artifact of the model.
Cyclone considers unique pointers ``consumed'' while a borrow exists, 
whereas Rust will allow the borrowed unique pointer to be used in certain situations
(namely, reading through an immutably borrowed unique pointer is allowed).
Cyclone supports abstraction over the aliasability and kind (box versus region) of a pointer,
but Rust does not currently support any such abstraction over pointer types.

Overall, Rust and Cyclone are each more expressive in different ways.
Cyclone's dynamic regions make it much easier to specify the lifetime of
a particular heap data structure, and they allow for heap structures that
are difficult or impossible to express in Rust, such as a doubly linked list.
Rust supports far more detailed reasoning about when memory should be freed
due to its flow-sensitivity, and Rust offers safer and more convenient
usage of linear data structures, particularly with regard to freeing them.
Cyclone still has much Rust can learn from, e.g. a mechanism similar to
dynamic regions could be used to support custom allocators in Rust.

@@

Boyland's alias burying \cite{aliasburying} has a similar goal to Rust, i.e.
using static checks rather than destructive reads, but differs in many details.
Boyland defines a strong version of the uniqueness invariant (true uniqueness) and a weak
version (allowing borrowed aliases), and then develops a system based on maintaing
the strong invariant. Rust meets only the weak uniqueness invariant, but it also
guarantees that aliased, mutable memory never exists, which allows for similar control
over mutation as the strong uniqueness invariant. The general idea of alias burying
is to allow for aliases, but to invalidate them when a unique field is read.
Rust ensures at a unique value will not be read if (mutable) aliases exist.
Rust's borrowed references are also more first-class than Boyland's, which cannot
be stored in data structures or returned from functions.
